---
title: "Ingesting hydrometric data to Bronze"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ingesting hydrometric data to Bronze}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

The Bronze tier is the immutable raw-data foundation of the reach.io store.
Every observation enters the pipeline here, stamped with provenance metadata
and written to a partitioned Parquet file. Nothing in Bronze is ever
overwritten — corrections and QC happen at Silver and Gold.

This vignette covers the full sequence from an empty disk to a populated
Bronze store:

```
1. setup_hydro_store()     — create the directory tree
2. build_gauge_registry()  — build the gauge lookup table
3. run_backfill()          — bulk ingest via the registry   (standard path)
   ingest_bulk_file()      — ingest a single bulk export    (alternative)
   download_hydrology()    — exploratory API download       (optional)
```

---

## 1. Create the directory structure

`setup_hydro_store()` builds the full Bronze / Silver / Gold / Register tree
in one call. It is safe to re-run — existing directories are left untouched.

```{r setup-store}
library(reach.io)

root <- "data/hydrometric"   # change to your chosen store root

setup_hydro_store(root)
```

```
  Created: data/hydrometric/bronze/hydrometric/EA/Q
  Created: data/hydrometric/bronze/hydrometric/EA/H
  Created: data/hydrometric/bronze/hydrometric/EA/P
  Created: data/hydrometric/bronze/hydrometric/WISKI/Q
  Created: data/hydrometric/bronze/hydrometric/WISKI/H
  Created: data/hydrometric/bronze/hydrometric/NRFA/Q
  Created: data/hydrometric/bronze/hydrometric/ratings
  Created: data/hydrometric/silver/hydrometric/Q
  ...
  Created: data/hydrometric/gold/hydrometric/ratings
  Created: data/hydrometric/register

Store setup complete under 'data/hydrometric':
  22 directories created, 0 already existed.
```

If you only need a subset of the default suppliers or data types, pass a
custom `categories` list:

```{r setup-store-custom}
# Hydrometric-only store; EA and WISKI sources; flow and level only
setup_hydro_store(
  root,
  categories = list(
    hydrometric = list(
      suppliers  = c("EA", "WISKI"),
      data_types = c("Q", "H")
    )
  )
)
```

---

## 2. Build the gauge registry

The registry (`gauge_registry.parquet`) is the single source of truth for
every gauge in the pipeline. It records the source system for each gauge so
that the router knows whether to call the EA Hydrology API (HDE), WISKI, or
read from a bulk export file.

Prepare a CSV with at minimum these columns:

| Column | Description | Example |
|---|---|---|
| `gauge_id` | Unique gauge identifier | `"39001"` |
| `source_system` | `"HDE"`, `"WISKI"`, `"BULK_FILE"`, or `"WISKI_ALL"` | `"HDE"` |
| `data_type` | `"flow"`, `"level"`, or `"rainfall"` | `"flow"` |
| `category` | `"hydrometric"`, `"radarH19"`, or `"MOSES"` | `"hydrometric"` |
| `catchment` | Catchment name (free text) | `"Thames at Kingston"` |
| `ea_site_ref` | EA station reference | `"39001"` |

```{r build-registry}
build_gauge_registry(
  input_csv   = "gauge_list.csv",       # your prepared CSV
  output_path = file.path(root, "register")
)
```

```
Validated 142 gauges.
  source_system breakdown:
    HDE       : 98
    WISKI     : 31
    BULK_FILE : 13

Written: data/hydrometric/register/gauge_registry.parquet
```

`build_gauge_registry()` adds three pipeline columns automatically:
`active` (default `TRUE`), `backfill_done` (default `FALSE`), and `notes`.
Set `active = FALSE` in your CSV to exclude tidal or decommissioned gauges
before they reach the router.

---

## 3. Standard path: registry-driven backfill

`run_backfill()` reads the registry, dispatches each active gauge to the
correct source system, writes Parquet to Bronze, and logs every outcome.
It runs gauges in parallel — one worker per available core by default.

```{r backfill}
run_backfill(
  registry_path = file.path(root, "register", "gauge_registry.parquet"),
  output_dir    = root,
  start_date    = "2015-01-01",
  end_date      = "2024-12-31",
  register_path = file.path(root, "register", "hydrometric_data_register.csv"),
  log_path      = "logs/backfill_2015_2024.csv"
)
```

```
Backfill started: 142 active gauges  |  8 workers  |  2015-01-01 to 2024-12-31
[========================================] 142/142 complete

  SUCCESS : 138 gauges  (97.2%)
  EMPTY   :   2 gauges  (1.4%)   — no data returned for date range
  FAILED  :   2 gauges  (1.4%)   — see logs/backfill_2015_2024.csv

Store populated: data/hydrometric/bronze/
```

Written files follow the partitioned layout
`bronze/<category>/<supplier>/<data_type>/<YYYY>/<dataset_id>.parquet`:

```
bronze/
└── hydrometric/
    ├── EA/
    │   ├── Q/
    │   │   ├── 2015/  EA_39001_Q_20250115.parquet
    │   │   ├── 2016/  EA_39001_Q_20250115.parquet
    │   │   └── ...
    │   └── H/
    │       └── ...
    └── WISKI/
        └── Q/
            └── ...
```

### Inspect the log

```{r inspect-log}
library(data.table)

log <- fread("logs/backfill_2015_2024.csv")

# Failed gauges with their error messages
log[status == "FAILED", .(gauge_id, error, elapsed_s)]

# Slow fetches (>60 s)
log[elapsed_s > 60, .(gauge_id, status, rows, elapsed_s)]
```

### Retry failed gauges

```{r retry}
run_backfill(
  registry_path = file.path(root, "register", "gauge_registry.parquet"),
  output_dir    = root,
  start_date    = "2015-01-01",
  end_date      = "2024-12-31",
  register_path = file.path(root, "register", "hydrometric_data_register.csv"),
  log_path      = "logs/backfill_2015_2024.csv",
  failed_only   = TRUE       # only gauges logged as FAILED previously
)
```

### Limit parallelism

On shared servers or when the EA API is under load, reduce the worker count:

```{r backfill-workers}
run_backfill(
  registry_path = file.path(root, "register", "gauge_registry.parquet"),
  output_dir    = root,
  start_date    = "2024-01-01",
  end_date      = "2024-12-31",
  register_path = file.path(root, "register", "hydrometric_data_register.csv"),
  log_path      = "logs/backfill_2024.csv",
  n_workers     = 4L
)
```

---

## 4. Alternative: ingest a single bulk export file

Use `ingest_bulk_file()` for one-off historical loads from a file already
on disk (e.g. a WISKI `.all` export or an EA bulk CSV), or for any gauge
not covered by the registry.

```{r ingest-bulk}
ingest_bulk_file(
  file_path     = "raw/EA_39001_flow_2000_2014.csv",
  site_id       = "39001",
  data_type     = "flow",
  output_dir    = root,
  register_path = file.path(root, "register", "hydrometric_data_register.csv"),
  supplier      = "Environment Agency",
  source_system = "BULK_FILE",
  received_by   = "J.Payne",
  notes         = "Pre-backfill historical load 2000–2014"
)
```

```
Parsed 525,960 rows from EA_39001_flow_2000_2014.csv
Written: data/hydrometric/bronze/hydrometric/EA/Q/2000/EA_39001_Q_20260319.parquet
         ...
         data/hydrometric/bronze/hydrometric/EA/Q/2014/EA_39001_Q_20260319.parquet
Register updated: data/hydrometric/register/hydrometric_data_register.csv
```

`ingest_bulk_file()` accepts `"csv"`, `"tsv"`, and `"fixed"` file formats
and automatically maps common column name variants (`Date`, `DateTime`,
`Value`, `Quality`, etc.) to the Bronze schema.

---

## 5. Optional: exploratory API download

Before committing to a full backfill, use `download_hydrology()` to pull a
short window into memory and inspect what the API returns.

```{r download-explore}
# Synchronous pull — data returned as a list of data.tables
result <- download_hydrology(
  parameters  = c("flow", "level"),
  from_date   = "2023-01-01",
  to_date     = "2023-03-31",
  method      = "sync",
  output      = "memory",
  period_name = "15min",
  wiski_ids   = "39001"
)

head(result$flow)
result$summary
```

Use `method = "batch"` for large requests — the API queues the job
asynchronously and `download_hydrology()` polls until it is ready:

```{r download-batch}
# Batch — suitable for multi-year or multi-gauge pulls
download_hydrology(
  parameters  = c("flow", "rainfall"),
  from_date   = "2010-01-01",
  to_date     = "2023-12-31",
  method      = "batch",
  output      = "disk",
  out_dir     = "scratch/api_pull",
  max_wait_s  = 900
)
```

---

## 6. Check the provenance register

Every ingested dataset — whether from a backfill, a bulk file, or a direct
API pull written to disk — creates a row in the register CSV:

```{r register}
register <- fread(
  file.path(root, "register", "hydrometric_data_register.csv")
)

# All flow datasets ingested this year
register[data_type == "Q" & date_received >= "2026-01-01",
         .(dataset_id, site_id, time_period_start, time_period_end,
           method_of_receipt, received_by)]
```

Key columns to know:

| Column | Description |
|---|---|
| `dataset_id` | Unique Bronze ID — join key to every Parquet row |
| `tier` | Always `"Bronze"` at this stage |
| `supplier` | Data provider |
| `site_id` | Gauge identifier |
| `data_type` | `Q`, `H`, or `P` |
| `time_period_start` / `_end` | Date range covered |
| `temporal_resolution` | `"15min"` or `"daily"` |
| `method_of_receipt` | `"API pull"`, `"bulk file"`, etc. |
| `received_by` | Custodian who ran the ingest |
| `file_path` | Full path to the Parquet on disk |

---

## Bronze schema

All Parquet files written to Bronze share this schema:

| Column | Type | Description |
|---|---|---|
| `timestamp` | POSIXct UTC | Observation datetime |
| `value` | float64 | Observed value (m³/s, mm, m) as received |
| `supplier_flag` | character | Quality code from supplier, or `NA` |
| `dataset_id` | character | Bronze dataset ID; join key to register |
| `site_id` | character | Supplier site identifier |
| `data_type` | character | `Q`, `H`, `P`, `SM`, or `SWE` |

No QC has been applied at this tier. `supplier_flag` carries whatever quality
code came from the source system. QC happens at Silver.
