---
title: "Data ingestion and exploration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data ingestion and exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

This vignette walks through the interactive data workflow in reach.io:
discovering stations, downloading flow, level, and rainfall data from the EA
Hydrology API, and exploring the results. It assumes the package is installed
and you have internet access to the EA API.

```
1. find_stations()        — discover stations and available measures
2. download_hydrology()   — download readings into memory as S7 objects
3. Inspect S7 objects     — understand the structure of returned data
4. Explore readings       — filter, reshape, and plot the data
5. Multiple gauges        — split a combined download by station
6. check_gaps()           — identify missing timesteps in the Bronze store
```

For bulk pipeline ingestion (registry + Parquet store) see
`vignette("bronze-ingestion")`.

---

## 1. Station discovery

Before downloading, use `find_stations()` to confirm which stations are
available and what measures each one carries. You can search by WISKI ID,
River Levels on the Internet (RLOIid), station name, notation, or geographic
proximity.

```{r find-by-wiski}
library(reach.io)

# Look up a single station by its WISKI ID
stns <- find_stations(wiski_ids = "SS92F014")
stns
#>    label    notation     wiskiID RLOIid     lat    long easting northing
#> 1: ...      <uuid>       SS92F014  5022   51.47  -0.31  ...     ...
#>    riverName   station.notation               parameter value_type period
#> 1: Thames      <uuid>-flow-i-900-m3s-...      flow      instantaneous  15min
#> 2: Thames      <uuid>-flow-m-86400-m3s-...    flow      mean           daily
#> 3: Thames      <uuid>-level-i-900-mASD-...    level     instantaneous  15min
```

The returned table has **one row per measure** — not one row per station.
The key columns are:

| Column | Description |
|---|---|
| `notation` | Station UUID — use this to filter `@readings` after download |
| `wiskiID` | WISKI identifier |
| `station.notation` | Full measure notation passed to the readings endpoint |
| `parameter` | `"flow"`, `"level"`, or `"rainfall"` |
| `value_type` | `"mean"`, `"instantaneous"`, `"total"`, etc. |
| `period` | `"daily"` or `"15min"` |

```{r find-multiple}
# Look up several stations at once — results are pooled and de-duplicated
stns <- find_stations(wiski_ids = c("SS92F014", "356337"))

# Fuzzy name search
thames <- find_stations(names = "Thames")

# By RLOIid
stns_rloi <- find_stations(rloi_ids = c("5022", "7001"))

# Proximity search — all stations within 15 km of a point
nearby <- find_stations(lat = 51.47, long = -0.31, dist = 15)

# Combine methods — any mix of identifiers is valid
combined <- find_stations(wiski_ids = "SS92F014",
                          lat = 51.47, long = -0.31, dist = 15)
```

To see only specific parameters after the lookup:

```{r filter-measures}
# All 15-minute instantaneous flow measures at these stations
stns[parameter == "flow" & period == "15min" & value_type == "instantaneous"]

# Available daily rainfall
stns[parameter == "rainfall" & period == "daily"]
```

---

## 2. Downloading data

`download_hydrology()` fetches readings for one or more stations and returns
typed S7 objects ready for analysis.

### 2.1 In-memory download

```{r download-basic}
result <- download_hydrology(
  parameters = c("flow", "rainfall"),
  from_date  = "2020-01-01",
  to_date    = "2023-12-31",
  wiski_ids  = c("SS92F014", "356337")
)
```

The return value is a named list — one element per requested parameter plus a
`$summary` table:

```{r result-names}
names(result)
#> [1] "flow"     "rainfall" "summary"
```

The summary gives a quick ingestion report:

```{r result-summary}
result$summary
#>    parameter  measure                                     file  rows    ok
#> 1: flow       <uuid>-flow-i-900-m3s-qualified              NA  140256  TRUE
#> 2: rainfall   <uuid>-rainfall-t-900-mm-qualified           NA  140256  TRUE
```

### 2.2 Choosing period and value type

The default period is **`"15min"`** for all parameters. Override with
`period_name` when daily aggregates are preferred:

```{r download-daily}
# Daily aggregates for flow and level
result_daily <- download_hydrology(
  parameters  = c("flow", "level"),
  from_date   = "2023-01-01",
  to_date     = "2023-12-31",
  period_name = "daily",
  wiski_ids   = "SS92F014"
)
```

`value_type` defaults to `"instantaneous"` for flow and level, and `"total"`
for rainfall (the only type available). Pass `"all"` to retrieve every
statistic for a parameter:

```{r download-all-vt}
# All daily level statistics (mean, min, max, instantaneous)
result_h <- download_hydrology(
  parameters  = "level",
  from_date   = "2023-01-01",
  to_date     = "2023-12-31",
  period_name = "daily",
  value_type  = "all",
  wiski_ids   = "SS92F014"
)
```

### 2.3 Saving to disk

For large date ranges or routine extracts, write directly to CSV files:

```{r download-disk}
download_hydrology(
  parameters       = c("flow", "level"),
  from_date        = "2015-01-01",
  to_date          = "2023-12-31",
  output           = "disk",
  out_dir          = "hydrology_data",
  observation_type = "Qualified",
  wiski_ids        = "SS92F014"
)
# Writes: hydrology_data/flow/*.csv, hydrology_data/level/*.csv
#         hydrology_data/download_summary.csv
```

---

## 3. Inspecting S7 objects

Each element of `result` is a typed S7 object. Printing one at the console
shows all available slots and the column names of `@readings`:

```{r print-object}
result$flow
#> <reach.io::Flow_15min>
#> Slots (@):
#>   @parameter      flow
#>   @period_name    15min
#>   @from_date      2020-01-01
#>   @to_date        2023-12-31
#>   @n_measures     1
#>   @n_rows         140,256
#>   @downloaded_at  2026-04-15 09:30:00 UTC
#>   @readings       data.table [140,256 × 5]
#>       $dateTime                POSIXct
#>       $date                    Date
#>       $value                   numeric
#>       $measure_notation        character
#>       $quality                 character
```

Access any slot with `@`:

```{r slot-access}
result$flow@parameter     # "flow"
result$flow@period_name   # "15min"
result$flow@n_measures    # 1
result$flow@n_rows        # 140256
result$flow@from_date     # "2020-01-01"
```

---

## 4. Exploring readings

`@readings` is a plain `data.table` — all standard `data.table` operations
apply directly.

```{r readings-head}
result$flow@readings
#>                    dateTime       date  value         measure_notation quality
#>    1: 2020-01-01 00:00:00 UTC 2020-01-01  12.34  <uuid>-flow-i-900-...      G
#>    2: 2020-01-01 00:15:00 UTC 2020-01-01  12.41  <uuid>-flow-i-900-...      G
#>    3: 2020-01-01 00:30:00 UTC 2020-01-01  12.38  <uuid>-flow-i-900-...      G
#>  ...
```

```{r readings-filter}
dt <- result$flow@readings

# Filter by date range
dt[date >= as.Date("2022-01-01") & date <= as.Date("2022-12-31")]

# Daily summary statistics from 15-minute data
dt[, .(mean_flow = mean(value, na.rm = TRUE),
       max_flow  = max(value,  na.rm = TRUE),
       n_obs     = .N),
   by = date]

# Check for missing values
dt[is.na(value)]
```

### 4.1 Extract as a plain data.table

```{r as-data-table}
# Equivalent to result$flow@readings but more explicit
dt <- as_data_table(result$flow)
```

### 4.2 Reshape to long format

`as_long()` prepends a `parameter` column, making it easy to bind different
parameter types into one table:

```{r as-long}
flow_long <- as_long(result$flow)
rain_long <- as_long(result$rainfall)

all_data <- rbind(flow_long, rain_long, fill = TRUE)
all_data[, .(rows = .N), by = parameter]
#>    parameter  rows
#> 1: flow       140256
#> 2: rainfall   140256
```

### 4.3 Basic plotting

For most plots it is easier to aggregate 15-minute data to daily resolution
first. With 15-minute data, merging directly by `date` would produce multiple
matches per day — always aggregate or merge by `dateTime` instead.

```{r plot-flow}
library(ggplot2)

# Aggregate to daily mean before plotting
dt_daily <- result$flow@readings[, .(value = mean(value, na.rm = TRUE)),
                                  by = date]

ggplot(dt_daily, aes(x = date, y = value)) +
  geom_line(colour = "#005EB8") +
  labs(
    title    = "Daily mean flow (aggregated from 15-minute)",
    subtitle = "SS92F014",
    x        = NULL,
    y        = "Flow (m\u00b3/s)"
  ) +
  theme_bw()
```

```{r plot-flow-rain}
# Aggregate both series to daily before merging
dt_flow <- result$flow@readings[,
             .(flow_m3s = mean(value, na.rm = TRUE)), by = date]
dt_rain <- result$rainfall@readings[,
             .(rain_mm  = sum(value,  na.rm = TRUE)), by = date]

combined <- merge(dt_flow, dt_rain, by = "date", all = TRUE)

rain_scale <- max(combined$flow_m3s, na.rm = TRUE) /
              max(combined$rain_mm,  na.rm = TRUE)

ggplot(combined, aes(x = date)) +
  geom_col(aes(y = rain_mm * rain_scale), fill = "#5A9FD4", alpha = 0.6) +
  geom_line(aes(y = flow_m3s), colour = "#005EB8", linewidth = 0.8) +
  scale_y_continuous(
    name     = "Flow (m\u00b3/s)",
    sec.axis = ggplot2::sec_axis(~ . / rain_scale, name = "Rainfall (mm)")
  ) +
  labs(title = "Daily flow and rainfall", x = NULL) +
  theme_bw()
```

---

## 5. Working with multiple gauges

When you download multiple WISKI IDs, all gauges for a parameter are combined
into **one S7 object**. The `measure_notation` column identifies which gauge
each row belongs to.

```{r multi-gauge-download}
result <- download_hydrology(
  parameters = c("flow", "rainfall"),
  from_date  = "2022-01-01",
  to_date    = "2024-12-31",
  wiski_ids  = c("SS92F014", "S11512_FW", "356337")
)

result$flow@n_measures     # 2 (one per flow gauge)
result$rainfall@n_measures # 1
```

### 5.1 Look up station UUIDs

`measure_notation` is a UUID-based EA string
(`<station-uuid>-flow-i-900-m3s-qualified`). The `notation` column returned
by `find_stations()` is the matching station UUID prefix.

```{r uuid-lookup}
stns <- find_stations(wiski_ids = c("SS92F014", "S11512_FW", "356337"))
stns[, .(wiskiID, notation, label)]
#>        wiskiID                               notation  label
#> 1:    SS92F014  47ed0d00-d4ab-4f3a-84c9-c9571d1a1cc1  ...
#> 2:   S11512_FW  a3b4c5d6-...                           ...
#> 3:      356337  f1e2d3c4-...                           ...

uuid_flow_a <- stns[wiskiID == "SS92F014",  notation[1L]]
uuid_flow_b <- stns[wiskiID == "S11512_FW", notation[1L]]
uuid_rain   <- stns[wiskiID == "356337",    notation[1L]]
```

### 5.2 Split into per-gauge S7 objects

```{r split-gauges}
flow_a <- Flow_15min(
  readings  = result$flow@readings[startsWith(measure_notation, uuid_flow_a)],
  from_date = result$flow@from_date,
  to_date   = result$flow@to_date
)

flow_b <- Flow_15min(
  readings  = result$flow@readings[startsWith(measure_notation, uuid_flow_b)],
  from_date = result$flow@from_date,
  to_date   = result$flow@to_date
)

rain <- Rainfall_15min(
  readings  = result$rainfall@readings[startsWith(measure_notation, uuid_rain)],
  from_date = result$rainfall@from_date,
  to_date   = result$rainfall@to_date
)

flow_a@n_rows  # rows for SS92F014 only
flow_b@n_rows  # rows for S11512_FW only
```

### 5.3 Compare gauges in one table

Aggregate to daily before comparing to avoid multiple rows per date:

```{r compare-gauges}
dt_a <- flow_a@readings[, .(flow_a = mean(value, na.rm = TRUE)), by = date]
dt_b <- flow_b@readings[, .(flow_b = mean(value, na.rm = TRUE)), by = date]

comparison <- merge(dt_a, dt_b, by = "date", all = TRUE)
comparison[, ratio := flow_a / flow_b]

# Correlation of daily means
cor(comparison$flow_a, comparison$flow_b, use = "complete.obs")
```

---

## 6. Annual summaries and hydrological year

`add_hydro_year()` appends a `hydro_year` integer and `hydro_doy`
(day-of-hydrological-year, starting 1 = 1 October) to any `data.table`
containing a `date` column.

```{r hydro-year}
dt <- result$flow@readings

dt <- reach.io::add_hydro_year(dt, start_month = 10L)

# Summarise at daily level first, then by hydro year
dt_daily <- dt[, .(mean_flow = mean(value, na.rm = TRUE),
                   hydro_year = hydro_year[1L]),
               by = date]

dt_daily[, .(
  mean_flow  = mean(mean_flow, na.rm = TRUE),
  total_days = data.table::uniqueN(date)
), by = hydro_year]
#>    hydro_year mean_flow total_days
#> 1:       2022     23.14        365
#> 2:       2023     18.92        365
#> 3:       2024     29.71        366
```

---

## 7. Double-mass plot

`double_mass_plot()` compares cumulative volumes on two axes — useful for
checking whether the flow record is consistent with the weighted rainfall
input.

```{r double-mass}
p <- double_mass_plot(
  x_data             = flow_a,
  y_data             = list(rain),
  catchment_area_km2 = 245.0,
  output_units       = "Ml",
  title              = "Flow vs areal rainfall — Thames tributary"
)
p

# Inspect the underlying cumulative series
attr(p, "data")
```

A straight line indicates a consistent ratio between flow and rainfall over the
period. Systematic breaks or kinks may indicate gauge issues, land-use change,
or abstractions not captured in the record.

---

## 8. Catalogue and gap detection

If data has been ingested to a Bronze store, the catalogue functions let you
query what is available without reading any Parquet files.

```{r catalogue}
# All gauges in the store
list_available_gauges(root = "data/hydrometric")

# Date coverage per gauge
summarise_coverage(root = "data/hydrometric", data_type = "Q")
#>    site_id  category     data_type  start       end         n_datasets
#> 1: SS92F014 hydrometric  Q         2010-01-01  2024-12-31   15
#> 2: S11512FW hydrometric  Q         2015-06-01  2024-12-31    9
```

`check_gaps()` reads the Parquet files for a gauge, deduplicates on
timestamp, and reports runs of missing observations:

```{r check-gaps}
check_gaps(
  root      = "data/hydrometric",
  site_id   = "SS92F014",
  data_type = "Q"
)
#> Gap analysis: SS92F014 | Q
#> Expected interval: 15min
#> ─────────────────────────────────────────────────────────────────────
#>    gap_start               gap_end                 n_missing
#> 1: 2018-03-01 06:00:00 UTC 2018-03-01 09:45:00 UTC 16
#> 2: 2020-11-15 12:00:00 UTC 2020-11-15 12:00:00 UTC  1
```

---

## Summary of key functions

| Task | Function |
|---|---|
| Discover stations | `find_stations(wiski_ids, rloi_ids, names, lat/long/dist)` |
| List measures | `get_measures(parameter, period_name, value_type)` |
| Download to memory | `download_hydrology(..., output = "memory")` |
| Download to disk | `download_hydrology(..., output = "disk", out_dir)` |
| Extract data.table | `as_data_table(obj)` or `obj@readings` |
| Long format | `as_long(obj)` |
| Add hydrological year | `add_hydro_year(dt, start_month)` |
| Double-mass plot | `double_mass_plot(x_data, y_data, catchment_area_km2)` |
| List Bronze gauges | `list_available_gauges(root)` |
| Date coverage | `summarise_coverage(root)` |
| Find data gaps | `check_gaps(root, site_id, data_type)` |
