---
title: "Data ingestion and exploration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data ingestion and exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

This vignette walks through the interactive data workflow in reach.io:
discovering stations, downloading flow, level, and rainfall data from the EA
Hydrology API, and exploring the results. It assumes the package is installed
and you have internet access to the EA API.

```
1. find_stations()        — discover stations and available measures
2. download_hydrology()   — download readings into memory as S7 objects
3. Inspect S7 objects     — understand the structure of returned data
4. Explore readings       — filter, reshape, and plot the data
5. Multiple gauges        — split a combined download by station
6. check_gaps()           — identify missing timesteps in the Bronze store
```

For bulk pipeline ingestion (registry + Parquet store) see
`vignette("bronze-ingestion")`.

---

## 1. Station discovery

Before downloading, use `find_stations()` to confirm which stations are
available and what measures each one carries. You can search by WISKI ID,
River Levels on the Internet (RLOIid), station name, notation, or geographic
proximity.

```{r find-by-wiski}
library(reach.io)

# Look up a single station by its WISKI ID
stns <- find_stations(wiski_ids = "SS92F014")
stns
#>    label    notation     wiskiID RLOIid     lat    long easting northing
#> 1: ...      <uuid>       SS92F014  5022   51.47  -0.31  ...     ...
#>    riverName   station.notation             parameter value_type period
#> 1: Thames      <uuid>-flow-m-...  flow       mean       daily
#> 2: Thames      <uuid>-flow-i-...  flow       instantaneous  15min
#> 3: Thames      <uuid>-level-i-... level      instantaneous  15min
```

The returned table has **one row per measure** — not one row per station.
The key columns are:

| Column | Description |
|---|---|
| `notation` | Station UUID — use this to filter `@readings` after download |
| `wiskiID` | WISKI identifier |
| `station.notation` | Full measure notation passed to the readings endpoint |
| `parameter` | `"flow"`, `"level"`, or `"rainfall"` |
| `value_type` | `"mean"`, `"instantaneous"`, `"total"`, etc. |
| `period` | `"daily"` or `"15min"` |

```{r find-multiple}
# Look up several stations at once — results are pooled and de-duplicated
stns <- find_stations(wiski_ids = c("SS92F014", "S11512_FW"))

# Fuzzy name search
thames <- find_stations(names = "Thames")

# By RLOIid
stns_rloi <- find_stations(rloi_ids = c("5022", "7001"))

# Proximity search — all stations within 15 km of a point
nearby <- find_stations(lat = 51.47, long = -0.31, dist = 15)

# Combine methods — any mix of identifiers is valid
combined <- find_stations(wiski_ids = "SS92F014",
                          lat = 51.47, long = -0.31, dist = 15)
```

To see only specific parameters after the lookup:

```{r filter-measures}
# All daily mean flow measures at these stations
stns[parameter == "flow" & period == "daily" & value_type == "mean"]

# Available 15-minute rainfall
stns[parameter == "rainfall" & period == "15min"]
```

---

## 2. Downloading data

`download_hydrology()` fetches readings for one or more stations and returns
typed S7 objects ready for analysis.

### 2.1 In-memory download

```{r download-basic}
result <- download_hydrology(
  parameters = c("flow", "rainfall"),
  from_date  = "2020-01-01",
  to_date    = "2023-12-31",
  wiski_ids  = c("SS92F014", "S11512_FW")
)
```

The return value is a named list — one element per requested parameter plus a
`$summary` table:

```{r result-names}
names(result)
#> [1] "flow"     "rainfall" "summary"
```

The summary gives a quick ingestion report:

```{r result-summary}
result$summary
#>    parameter  measure                                      file rows  ok
#> 1: flow       <uuid>-flow-m-86400-m3s-qualified            NA   1461  TRUE
#> 2: flow       <uuid>-flow-m-86400-m3s-qualified            NA   1461  TRUE
#> 3: rainfall   <uuid>-rainfall-t-86400-mm-qualified         NA   1461  TRUE
#> 4: rainfall   <uuid>-rainfall-t-86400-mm-qualified         NA   1461  TRUE
```

### 2.2 Choosing period and value type

The default period is `"daily"` for all parameters. Override with
`period_name`:

```{r download-15min}
# 15-minute flow and level
result_15 <- download_hydrology(
  parameters  = c("flow", "level"),
  from_date   = "2023-01-01",
  to_date     = "2023-12-31",
  period_name = "15min",
  wiski_ids   = "SS92F014"
)
```

`value_type` defaults to `"mean"` for flow, `"total"` for rainfall, and
`"instantaneous"` for level. Pass `"all"` to retrieve every statistic for a
parameter:

```{r download-all-vt}
# All level statistics (min, max, instantaneous)
result_h <- download_hydrology(
  parameters = "level",
  from_date  = "2023-01-01",
  to_date    = "2023-12-31",
  value_type = "all",
  wiski_ids  = "SS92F014"
)
```

### 2.3 Saving to disk

For large date ranges or routine extracts, write directly to CSV files:

```{r download-disk}
download_hydrology(
  parameters       = c("flow", "level"),
  from_date        = "2015-01-01",
  to_date          = "2023-12-31",
  output           = "disk",
  out_dir          = "hydrology_data",
  observation_type = "Qualified",
  wiski_ids        = "SS92F014"
)
# Writes: hydrology_data/flow/*.csv, hydrology_data/level/*.csv
#         hydrology_data/download_summary.csv
```

---

## 3. Inspecting S7 objects

Each element of `result` is a typed S7 object. Printing one at the console
shows all available slots and the column names of `@readings`:

```{r print-object}
result$flow
#> <reach.io::Flow_Daily>
#> Slots (@):
#>   @parameter      flow
#>   @period_name    daily
#>   @from_date      2020-01-01
#>   @to_date        2023-12-31
#>   @n_measures     2
#>   @n_rows         2,922
#>   @downloaded_at  2026-04-15 09:30:00 UTC
#>   @readings       data.table [2,922 × 5]
#>       $dateTime                POSIXct
#>       $date                    Date
#>       $value                   numeric
#>       $measure_notation        character
#>       $quality                 character
```

Access any slot with `@`:

```{r slot-access}
result$flow@parameter     # "flow"
result$flow@period_name   # "daily"
result$flow@n_measures    # 2
result$flow@n_rows        # 2922
result$flow@from_date     # "2020-01-01"
```

---

## 4. Exploring readings

`@readings` is a plain `data.table` — all standard `data.table` operations
apply directly.

```{r readings-head}
result$flow@readings
#>              dateTime       date  value         measure_notation quality
#>    1: 2020-01-01 UTC 2020-01-01  12.34  <uuid>-flow-m-86400-...      G
#>    2: 2020-01-02 UTC 2020-01-02  14.21  <uuid>-flow-m-86400-...      G
#>  ...
```

```{r readings-filter}
dt <- result$flow@readings

# Filter by date
dt[date >= as.Date("2022-01-01") & date <= as.Date("2022-12-31")]

# Summary statistics
dt[, .(mean_flow = mean(value, na.rm = TRUE),
       max_flow  = max(value,  na.rm = TRUE),
       n_days    = .N),
   by = data.table::year(date)]

# Check for missing values
dt[is.na(value)]
```

### 4.1 Extract as a plain data.table

```{r as-data-table}
# Equivalent to result$flow@readings but more explicit
dt <- as_data_table(result$flow)
```

### 4.2 Reshape to long format

`as_long()` prepends a `parameter` column, making it easy to bind different
parameter types into one table:

```{r as-long}
flow_long <- as_long(result$flow)
rain_long <- as_long(result$rainfall)

all_data <- rbind(flow_long, rain_long, fill = TRUE)
all_data[, .(rows = .N), by = parameter]
#>    parameter  rows
#> 1: flow       2922
#> 2: rainfall   2922
```

### 4.3 Basic plotting

```{r plot-flow}
library(ggplot2)

dt <- result$flow@readings

ggplot(dt, aes(x = date, y = value)) +
  geom_line(colour = "#005EB8") +
  labs(
    title    = "Daily mean flow",
    subtitle = "SS92F014",
    x        = NULL,
    y        = "Flow (m\u00b3/s)"
  ) +
  theme_bw()
```

```{r plot-flow-rain}
# Flow and rainfall together (dual axis)
dt_flow <- result$flow@readings[, .(date, flow_m3s = value)]
dt_rain <- result$rainfall@readings[, .(date, rain_mm = value)]

combined <- merge(dt_flow, dt_rain, by = "date", all = TRUE)

rain_scale <- max(combined$flow_m3s, na.rm = TRUE) /
              max(combined$rain_mm,  na.rm = TRUE)

ggplot(combined, aes(x = date)) +
  geom_col(aes(y = rain_mm * rain_scale), fill = "#5A9FD4", alpha = 0.6) +
  geom_line(aes(y = flow_m3s), colour = "#005EB8", linewidth = 0.8) +
  scale_y_continuous(
    name     = "Flow (m\u00b3/s)",
    sec.axis = ggplot2::sec_axis(~ . / rain_scale, name = "Rainfall (mm)")
  ) +
  labs(title = "Daily flow and rainfall", x = NULL) +
  theme_bw()
```

---

## 5. Working with multiple gauges

When you download multiple WISKI IDs, all gauges for a parameter are combined
into **one S7 object**. The `measure_notation` column identifies which gauge
each row belongs to.

```{r multi-gauge-download}
result <- download_hydrology(
  parameters = c("flow", "rainfall"),
  from_date  = "2022-01-01",
  to_date    = "2024-12-31",
  wiski_ids  = c("SS92F014", "S11512_FW", "RAIN001")
)

result$flow@n_measures  # 2 (one per flow gauge)
```

### 5.1 Look up station UUIDs

`measure_notation` is a UUID-based EA string
(`<station-uuid>-flow-m-86400-m3s-qualified`). The `notation` column returned
by `find_stations()` is the matching station UUID prefix.

```{r uuid-lookup}
stns <- find_stations(wiski_ids = c("SS92F014", "S11512_FW", "RAIN001"))
stns[, .(wiskiID, notation, label)]
#>        wiskiID                               notation  label
#> 1:    SS92F014  47ed0d00-d4ab-4f3a-84c9-c9571d1a1cc1  ...
#> 2:   S11512_FW  a3b4c5d6-...                           ...
#> 3:     RAIN001  f1e2d3c4-...                           ...

uuid_flow_a <- stns[wiskiID == "SS92F014",  notation[1L]]
uuid_flow_b <- stns[wiskiID == "S11512_FW", notation[1L]]
uuid_rain   <- stns[wiskiID == "RAIN001",   notation[1L]]
```

### 5.2 Split into per-gauge S7 objects

```{r split-gauges}
flow_a <- Flow_Daily(
  readings  = result$flow@readings[startsWith(measure_notation, uuid_flow_a)],
  from_date = result$flow@from_date,
  to_date   = result$flow@to_date
)

flow_b <- Flow_Daily(
  readings  = result$flow@readings[startsWith(measure_notation, uuid_flow_b)],
  from_date = result$flow@from_date,
  to_date   = result$flow@to_date
)

rain <- Rainfall_Daily(
  readings  = result$rainfall@readings[startsWith(measure_notation, uuid_rain)],
  from_date = result$rainfall@from_date,
  to_date   = result$rainfall@to_date
)

flow_a@n_rows  # rows for SS92F014 only
flow_b@n_rows  # rows for S11512_FW only
```

### 5.3 Compare gauges in one table

```{r compare-gauges}
dt_a <- flow_a@readings[, .(date, flow_a = value)]
dt_b <- flow_b@readings[, .(date, flow_b = value)]

comparison <- merge(dt_a, dt_b, by = "date", all = TRUE)
comparison[, ratio := flow_a / flow_b]

# Correlation
cor(comparison$flow_a, comparison$flow_b, use = "complete.obs")
```

---

## 6. Annual summaries and hydrological year

`add_hydro_year()` appends a `hydro_year` integer and `hydro_doy`
(day-of-hydrological-year, starting 1 = 1 October) to any `data.table`
containing a `date` column.

```{r hydro-year}
dt <- result$flow@readings

dt <- reach.io::add_hydro_year(dt, start_month = 10L)

dt[, .(
  mean_flow  = mean(value,  na.rm = TRUE),
  total_days = .N
), by = hydro_year]
#>    hydro_year mean_flow total_days
#> 1:       2022     23.14        365
#> 2:       2023     18.92        365
#> 3:       2024     29.71        366
```

---

## 7. Double-mass plot

`double_mass_plot()` compares cumulative volumes on two axes — useful for
checking whether the flow record is consistent with the weighted rainfall
input.

```{r double-mass}
p <- double_mass_plot(
  x_data             = flow_a,
  y_data             = list(rain),
  catchment_area_km2 = 245.0,
  output_units       = "Ml",
  title              = "Flow vs areal rainfall — Thames tributary"
)
p

# Inspect the underlying cumulative series
attr(p, "data")
```

A straight line indicates a consistent ratio between flow and rainfall over the
period. Systematic breaks or kinks may indicate gauge issues, land-use change,
or abstractions not captured in the record.

---

## 8. Catalogue and gap detection

If data has been ingested to a Bronze store, the catalogue functions let you
query what is available without reading any Parquet files.

```{r catalogue}
# All gauges in the store
list_available_gauges(root = "data/hydrometric")

# Date coverage per gauge
summarise_coverage(root = "data/hydrometric", data_type = "Q")
#>    site_id  category     data_type  start       end         n_datasets
#> 1: SS92F014 hydrometric  Q         2010-01-01  2024-12-31   15
#> 2: S11512FW hydrometric  Q         2015-06-01  2024-12-31    9
```

`check_gaps()` reads the Parquet files for a gauge, deduplicates on
timestamp, and reports runs of missing observations:

```{r check-gaps}
check_gaps(
  root    = "data/hydrometric",
  site_id = "SS92F014",
  data_type = "Q"
)
#> Gap analysis: SS92F014 | Q
#> Expected interval: daily
#> ─────────────────────────────────────────────────────────────────────
#>    gap_start   gap_end    n_missing
#> 1: 2018-03-01  2018-03-04  3
#> 2: 2020-11-15  2020-11-15  1
```

---

## Summary of key functions

| Task | Function |
|---|---|
| Discover stations | `find_stations(wiski_ids, rloi_ids, names, lat/long/dist)` |
| List measures | `get_measures(parameter, period_name, value_type)` |
| Download to memory | `download_hydrology(..., output = "memory")` |
| Download to disk | `download_hydrology(..., output = "disk", out_dir)` |
| Extract data.table | `as_data_table(obj)` or `obj@readings` |
| Long format | `as_long(obj)` |
| Add hydrological year | `add_hydro_year(dt, start_month)` |
| Double-mass plot | `double_mass_plot(x_data, y_data, catchment_area_km2)` |
| List Bronze gauges | `list_available_gauges(root)` |
| Date coverage | `summarise_coverage(root)` |
| Find data gaps | `check_gaps(root, site_id, data_type)` |
