---
title: "Rating curve lifecycle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rating curve lifecycle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

A rating curve translates gauge stage (water level, m) into discharge (flow,
m³/s) using the equation:

$$Q = C \cdot (h - a)^b$$

where $C$ is a multiplier, $a$ is a datum offset, and $b$ is an exponent.
Real-world curves are split into multiple **limbs**, each covering a different
stage range (e.g. low in-channel flow, bankfull, out-of-bank). The upper limbs
are often extrapolated beyond measured gaugings and are flagged as **doubtful**.

Ratings change over time as channels shift, structures are modified, or
resurveys are carried out. reach.io tracks this through:

- **`RatingCurve`** — a single set of limbs with an optional validity window.
- **`RatingSet`** — an ordered collection of `RatingCurve` objects covering
  non-overlapping periods.

Storage follows the same Bronze → Silver → Gold medallion tier model used for
observation data.

```
Bronze   As-received from source (WISKI export, consultant report).
         Immutable once saved.

Silver   QC-reviewed: continuity checked, limb boundaries verified.
         Not yet formally approved.

Gold     Approved for production use. Default tier for apply_rating().
```

---

## 1. Building a rating curve

Construct a `RatingCurve` from a `data.table` of limb parameters. The
constructor accepts common column name variants (`Lower.level`, `multiplier`,
`Exponent`, etc.) so real-world exports rarely need pre-processing.

```{r build}
library(reach.io)
library(data.table)

limbs <- data.table(
  lower    = c(0.000, 0.095, 2.000),
  upper    = c(0.095, 2.000,   Inf),
  C        = c(12.2495, 23.252, 23.252),
  a        = c(-0.003, -0.098,  -0.098),
  b        = c( 1.641,  2.706,   2.706),
  doubtful = c(FALSE,   FALSE,   TRUE)   # upper limb = out-of-bank
)

rc <- RatingCurve(
  limbs      = limbs,
  valid_from = as.Date("2019-10-01"),
  station_id = "510310",
  source     = "WISKI"
)

rc
```

```
<RatingCurve>
  Station:  510310
  Source:   WISKI
  Valid:    2019-10-01 to open
  Limbs:    3

    #   Lower   Upper         C         a         b  Doubtful
    1   0.000   0.095   12.2495    -0.003     1.641     FALSE
    2   0.095   2.000   23.252    -0.098      2.706     FALSE
    3   2.000     Inf   23.252    -0.098      2.706      TRUE
```

---

## 2. Checking limb continuity

At each boundary between limbs the rating equation should produce the same $Q$
from both sides. In practice, curves fitted independently to separate gaugings
often have a small step discontinuity.

```{r check}
check_limb_continuity(rc)
```

```
2 of 2 junction(s) exceed the 1% continuity tolerance.
   junction h_junction Q_lower_limb Q_upper_limb  gap_m3s  gap_pct
      1-2       0.095       0.0412       0.0389  -0.0023     5.6%
      2-3       2.000       1.3241       1.3241   0.0000     0.0%
```

`check_limb_continuity()` returns its result invisibly so you can capture and
inspect it:

```{r check-capture}
continuity <- check_limb_continuity(rc, tol_pct = 1)
continuity[connected == FALSE]
```

---

## 3. Correcting continuity

`fix_limb_continuity()` adjusts the $C$ parameter of each limb (propagating
upward from the anchor limb) so that $Q$ is exactly continuous at every
junction. The shape parameters $a$ and $b$ are never changed.

```{r fix}
rc_fixed <- fix_limb_continuity(rc)
```

```
Adjusting C parameters (anchor = limb 1, propagating upward):
  Limb 1 -> 2  h = 0.095 m:  C  23.252  ->  22.874  (-1.63%)
  Limb 2 -> 3  h = 2.000 m:  C  unchanged
```

Verify the result:

```{r verify}
check_limb_continuity(rc_fixed)
# All 2 junction(s) connect within 1%.
```

---

## 4. Saving to Bronze

Save the original, unmodified curve as-received. Bronze is the immutable record
of what was supplied.

```{r save-bronze}
root <- "data/hydro"   # your store root

save_rating(
  rc,
  root     = root,
  tier     = "bronze",
  saved_by = "J.Payne",
  notes    = "WISKI export 2023-11-01, version 3.2"
)
# Saved RatingCurve [bronze] v1 for site '510310'
#   -> data/hydro/bronze/hydrometric/ratings/ratings.parquet
```

---

## 5. Promoting to Silver

After review, save the QC-corrected curve to Silver.

```{r save-silver}
save_rating(
  rc_fixed,
  root     = root,
  tier     = "silver",
  saved_by = "J.Payne",
  notes    = "Limb 2 C corrected for continuity with fix_limb_continuity()"
)
# Saved RatingCurve [silver] v1 for site '510310'
```

---

## 6. Approving to Gold

Once formally signed off, promote to Gold. This is the version that
`apply_rating()` will use in production.

```{r save-gold}
save_rating(
  rc_fixed,
  root     = root,
  tier     = "gold",
  saved_by = "J.Payne",
  notes    = "Approved by lead hydrologist 2023-11-15"
)
# Saved RatingCurve [gold] v1 for site '510310'
```

The governance register now has one row per tier for this version:

```{r list}
list_ratings(root, site_id = "510310")
```

```
<Rating register>  3 version(s)

   site_id   tier version valid_from valid_to n_limbs source saved_by   saved_at
1.  510310 bronze       1 2019-10-01       NA       3  WISKI  J.Payne 2023-11-01
2.  510310 silver       1 2019-10-01       NA       3  WISKI  J.Payne 2023-11-01
3.  510310   gold       1 2019-10-01       NA       3  WISKI  J.Payne 2023-11-15
```

---

## 7. Applying to level data

Load the approved curve and apply it to a `Level_Daily` or `Level_15min`
object. The output is a `Flow_Daily` or `Flow_15min` with an additional
`doubtful` column flagging flows derived from extrapolated limbs.

```{r apply}
rc_gold <- load_rating("510310", root = root)   # returns RatingSet
flow    <- apply_rating(level_obj, rc_gold)

# Inspect
head(flow@readings)
#             dateTime       date  value measure_notation doubtful
# 1: 2019-10-01 00:00:00 2019-10-01  0.412     rated_flow    FALSE
# 2: 2019-10-01 00:15:00 2019-10-01  0.418     rated_flow    FALSE
# 3: 2019-10-01 00:30:00 2019-10-01  1.853     rated_flow    FALSE
# 4: 2019-10-01 00:45:00 2019-10-01  3.421     rated_flow     TRUE  # out-of-bank

# Flows derived from the doubtful (out-of-bank) limb
flow@readings[doubtful == TRUE]
```

---

## 8. Time-varying ratings

When the channel has changed — due to a flood event, maintenance works, or a
resurvey — create a new `RatingCurve` for the new period and combine them into
a `RatingSet`. `apply_rating()` automatically selects the correct curve for
each observation date.

```{r rating-set}
# Original curve: valid 2019-10-01 to 2022-06-30
rc_v1 <- RatingCurve(
  limbs      = limbs_v1,
  valid_from = as.Date("2019-10-01"),
  valid_to   = as.Date("2022-06-30"),
  station_id = "510310",
  source     = "WISKI"
)

# Post-survey curve: valid from 2022-07-01 onwards
rc_v2 <- RatingCurve(
  limbs      = limbs_v2,
  valid_from = as.Date("2022-07-01"),
  station_id = "510310",
  source     = "WISKI"
)

rs <- RatingSet(curves = list(rc_v1, rc_v2), station_id = "510310")

# Save both to Gold in one call
save_rating(rs, root = root, tier = "gold", saved_by = "J.Payne")

# Load and apply — each observation uses the curve valid on its date
rs_gold <- load_rating("510310", root = root)
flow    <- apply_rating(level_obj, rs_gold)
```

Observations outside all validity windows are rated as `NA` with a warning.

---

## 9. Comparing tiers

Load from different tiers to compare what changed between the raw and corrected
curves:

```{r compare}
rc_bronze <- load_rating("510310", root, tier = "bronze", version = 1)
rc_gold   <- load_rating("510310", root, tier = "gold",   version = 1)

# Side-by-side continuity check
check_limb_continuity(rc_bronze)   # junction 1-2 has a gap
check_limb_continuity(rc_gold)     # all connected

# Gold register entries only
list_ratings(root, tier = "gold")
```

---

## Directory layout

`setup_hydro_store()` creates the full tree including the ratings directories:

```
<root>/
├── bronze/hydrometric/ratings/ratings.parquet
├── silver/hydrometric/ratings/ratings.parquet
├── gold/hydrometric/ratings/ratings.parquet
└── register/rating_register.csv
```

Each Parquet file holds all sites and all versions at that tier, with columns
`site_id` and `version` as the composite key. The register CSV is the single
audit trail across all three tiers.
