---
title: "Silver tier: quality control and data access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Silver tier: quality control and data access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Note:** Silver tier QC is currently trial / skeleton code. Thresholds and
> check logic have not been validated against EA operational data and will
> change. Do not use in production workflows yet.

## Overview

The Silver tier applies automated quality control to Bronze Parquet data,
adding QC flags and a QC-accepted value column. It sits between raw Bronze
ingestion and the Gold tier, which will hold derived and approved outputs.

```
Bronze  →  promote_to_silver()  →  Silver  →  read_silver()  →  analysis
  raw            QC flags             flagged        filtered
```

The Silver schema extends Bronze by four columns:

| Column | Type | Description |
|---|---|---|
| `qc_flag` | integer | 1 Good · 2 Estimated · 3 Suspect · 4 Rejected |
| `qc_value` | float64 | Accepted reading; `NA` where Rejected |
| `qc_y_code` / `qc_h_code` / `qc_p_code` | integer | Data-type-specific anomaly code |
| `qc_flagged_at` | POSIXct | Timestamp of the promotion run |

Flag 2 (Estimated) is **never assigned automatically** — it is reserved for
values that have been manually corrected or gap-filled downstream.

---

## 1. Promoting Bronze data to Silver

`promote_to_silver()` reads a Bronze Parquet file (or a `data.table`
conforming to the Bronze schema), applies QC checks appropriate to the
data type, and writes the result to the Silver store.

```{r promote-basic}
library(reach.io)

# Promote a single Bronze file
result <- promote_to_silver(
  bronze_data = "data/hydrometric/bronze/hydrometric/EA/Q/2023/EA_39001_Q_20260115.parquet",
  output_dir  = "data/hydrometric"
)

result
#> <data.table [35040 × 10]>
#>    timestamp           value supplier_flag dataset_id       site_id data_type
#>    qc_flag qc_value qc_y_code qc_flagged_at
```

The function returns the Silver `data.table` invisibly when writing to disk
(`write_output = TRUE`, the default). Assign it to inspect the result in
memory.

### 1.1 Promoting all Bronze files for a gauge

```{r promote-loop}
# Locate all Bronze files for site 39001, data type Q
bronze_files <- list.files(
  path       = "data/hydrometric/bronze/hydrometric/EA/Q",
  pattern    = "^EA_39001_Q_.*\\.parquet$",
  full.names = TRUE,
  recursive  = TRUE
)

for (f in bronze_files) {
  message("Promoting: ", basename(f))
  promote_to_silver(f, output_dir = "data/hydrometric")
}
```

### 1.2 Inspecting QC results before writing

Set `write_output = FALSE` to check results without writing Parquet files:

```{r promote-dryrun}
result <- promote_to_silver(
  bronze_data  = "data/hydrometric/bronze/hydrometric/EA/Q/2023/EA_39001_Q_20260115.parquet",
  output_dir   = "data/hydrometric",
  write_output = FALSE
)

# QC flag breakdown
result[, .N, by = qc_flag][order(qc_flag)]
#>    qc_flag     N
#> 1:       1 34891
#> 2:       3   127
#> 3:       4    22

# What anomaly codes were assigned?
result[qc_flag > 1L, .N, by = qc_y_code][order(qc_y_code)]
#>    qc_y_code   N
#> 1:         2  88   # relative spike (suspect)
#> 2:         3  39   # absolute spike (suspect)
#> 3:         8  22   # combined spike (rejected)
```

---

## 2. QC checks by data type

Checks are applied automatically based on the `data_type` column in the
Bronze data. No configuration is needed for the defaults.

### 2.1 Flow (Q) — Y-digit codes

| Y-code | Anomaly | Default flag |
|---|---|---|
| 0 | Clean | Good (1) |
| 1 | Negative value | Rejected (4) |
| 2 | Relative spike (≥10× previous) | Suspect (3) |
| 3 | Absolute spike (MAD-based) | Suspect (3) |
| 4 | Drop (≥90% in one step) | Suspect (3) |
| 5 | Rapid fluctuation (alternating) | Suspect (3) |
| 6 | Truncated low (stuck at sensor floor) | Suspect (3) |
| 7 | Truncated high (stuck at sensor cap) | Suspect (3) |
| 8 | Combined spike (codes 2 + 3) | Rejected (4) |
| 9 | Combined drop + negative/truncation | Rejected (4) |

Ultrasonic and tidal gauges can legitimately record negative values. Disable
the negative-flow check with `allow_negative = TRUE`:

```{r promote-tidal}
promote_to_silver(
  bronze_data    = "data/hydrometric/bronze/hydrometric/EA/Q/2023/EA_tidal_Q_20260115.parquet",
  output_dir     = "data/hydrometric",
  allow_negative = TRUE
)
```

Adjusting spike sensitivity (e.g., for a naturally flashy catchment):

```{r promote-flashy}
promote_to_silver(
  bronze_data          = "data/hydrometric/bronze/hydrometric/EA/Q/2023/EA_39001_Q_20260115.parquet",
  output_dir           = "data/hydrometric",
  relative_spike_ratio = 20,   # require 20× increase before flagging
  absolute_spike_k     = 8     # 8 MADs rather than 5
)
```

### 2.2 Stage / level (H) — H-digit codes

| H-code | Anomaly | Default flag |
|---|---|---|
| 0 | Clean | Good (1) |
| 1 | Out-of-range (below datum / above credible max) | Suspect (3) |
| 2 | Spike-and-return (debris, ice contact) | Suspect (3) |
| 3 | Implausible rate of change (>0.5 m per step) | Suspect (3) |
| 4 | Flatline (≥4 identical consecutive values) | Suspect (3) |
| 5 | Datum shift (persistent offset between windows) | Suspect (3) |

All H anomalies produce Suspect (3), not Rejected — stage is noisier than
flow and manual review is expected before rejection.

Setting datum bounds for a specific gauge:

```{r promote-level}
promote_to_silver(
  bronze_data  = "data/hydrometric/bronze/hydrometric/EA/H/2023/EA_39001_H_20260115.parquet",
  output_dir   = "data/hydrometric",
  min_datum    = -2.0,    # below -2 mAOD is physically implausible
  max_credible = 8.5      # above 8.5 mAOD is above bank-full
)
```

### 2.3 Rainfall (P) — P-digit codes

| P-code | Anomaly | Default flag |
|---|---|---|
| 0 | Clean | Good (1) |
| 1 | Negative value | Rejected (4) |
| 2 | Intensity spike (>50 mm/15min) | Suspect (3) |
| 3 | Daily cap exceeded (>300 mm/24hr) | Suspect (3) |
| 4 | Dry-run during wet period | Suspect (3) |
| 5 | Tipping-bucket reset spike | Rejected (4) |

```{r promote-rainfall}
promote_to_silver(
  bronze_data         = "data/hydrometric/bronze/hydrometric/EA/P/2023/EA_39001_P_20260115.parquet",
  output_dir          = "data/hydrometric",
  max_intensity_15min = 30,   # lower threshold for typical upland gauge
  max_daily_mm        = 200
)
```

---

## 3. Standalone QC checks

Three checks require additional context (a paired series or neighbouring
gauge data) and so are not called automatically inside `promote_to_silver()`.
They can be applied to the Silver `data.table` after promotion.

### 3.1 Stage–flow consistency

Flags periods where stage and flow move in sustained opposing directions
(e.g., rising stage but falling flow, suggesting a blockage or rating error):

```{r stage-flow}
silver_q <- promote_to_silver(
  "data/hydrometric/bronze/hydrometric/EA/Q/2023/EA_39001_Q_20260115.parquet",
  output_dir   = "data/hydrometric",
  write_output = FALSE
)
silver_h <- promote_to_silver(
  "data/hydrometric/bronze/hydrometric/EA/H/2023/EA_39001_H_20260115.parquet",
  output_dir   = "data/hydrometric",
  write_output = FALSE
)

# Align on timestamp then apply the check
paired <- merge(
  silver_q[, .(timestamp, flow  = qc_value)],
  silver_h[, .(timestamp, stage = qc_value)],
  by = "timestamp"
)

inconsistent <- qc_h_stage_flow_consistency(
  stage = paired$stage,
  flow  = paired$flow,
  consistency_window = 4L   # 4 consecutive opposing steps required
)

sum(inconsistent, na.rm = TRUE)  # number of flagged timesteps
```

### 3.2 Rainfall dry-run detection

Flags extended zero-rainfall periods at a gauge when nearby gauges are
recording rain, suggesting a blocked or frozen tipping bucket:

```{r dry-run}
# Build a wet-period mask from a reference (neighbouring) gauge
silver_ref <- arrow::read_parquet("data/hydrometric/silver/hydrometric/P/2023/EA_REF_P_20260115.parquet")
wet_mask   <- silver_ref$qc_value > 0.2  # reference records > 0.2 mm = "wet"

flagged <- qc_p_dry_run(
  value         = silver_p$qc_value,
  wet_period_mask = wet_mask,
  dry_run_min_steps = 12L  # 3 hours of zeros during wet = suspicious
)
```

### 3.3 Spatial consistency (temporal isolation)

Flags rainfall that is isolated from neighbouring gauges — high at one
gauge while all neighbours are dry:

```{r spatial-consistency}
# Build a neighbour matrix: one column per gauge, rows = timesteps
neighbour_values <- cbind(
  gauge_a = silver_a$qc_value,
  gauge_b = silver_b$qc_value,
  gauge_c = silver_c$qc_value
)

flagged <- qc_p_temporal_consistency(
  value                = silver_p$qc_value,
  neighbours           = neighbour_values,
  isolation_threshold  = 0.2,   # > 0.2 mm counts as "wet"
  isolation_fraction   = 0.8    # flag if ≥80% of neighbours are dry
)
```

---

## 4. Reading Silver data

`read_silver()` locates Silver Parquet files, applies date and quality
filters, and returns a typed S7 object consistent with `download_hydrology()`
output.

```{r read-silver-basic}
# Read Good-only flow data for 2023
flow_obj <- read_silver(
  root      = "data/hydrometric",
  gauge_id  = "39001",
  data_type = "Q",
  start     = "2023-01-01",
  end       = "2023-12-31"
)

flow_obj
#> <reach.io::Flow_15min>
#> Slots (@):
#>   @parameter      flow
#>   @period_name    15min
#>   @from_date      2023-01-01
#>   @to_date        2023-12-31
#>   @n_measures     1
#>   @n_rows         34,891
#>   @downloaded_at  2026-04-15 10:00:00 UTC
#>   @readings       data.table [34,891 × 10]
#>       $dateTime                POSIXct
#>       $date                    Date
#>       $value                   numeric      ← qc_value (NA = Rejected)
#>       $measure_notation        character    ← dataset_id
#>       $raw_value               numeric      ← original Bronze value
#>       $supplier_flag           character
#>       $site_id                 character
#>       $qc_flag                 integer
#>       $qc_y_code               integer
#>       $qc_flagged_at           POSIXct
```

The `value` slot in `@readings` holds `qc_value` — the QC-accepted reading.
The original Bronze value is always present as `raw_value` for transparency.

### 4.1 Quality thresholds

Control which flags are included via `min_quality`:

```{r read-quality}
# Good only (default) — excludes Suspect and Rejected
good_only <- read_silver("data/hydrometric", "39001", "Q",
                         min_quality = 1L)

# Good + Suspect — useful for exploratory analysis
incl_suspect <- read_silver("data/hydrometric", "39001", "Q",
                            min_quality = 3L)

# All readings including Rejected (raw_value always non-NA)
all_flags <- read_silver("data/hydrometric", "39001", "Q",
                         min_quality = NULL)

# Compare coverage
good_only@n_rows     # 34891
incl_suspect@n_rows  # 35018
all_flags@n_rows     # 35040
```

### 4.2 Inspecting flag distributions

```{r flag-inspection}
dt <- read_silver("data/hydrometric", "39001", "Q",
                  min_quality = NULL)@readings

# Count by flag
dt[, .N, by = qc_flag][order(qc_flag)]
#>    qc_flag     N
#> 1:       1 34891
#> 2:       3   127
#> 3:       4    22

# Percentage good
dt[, .(pct_good = round(mean(qc_flag == 1L) * 100, 1))]
#>    pct_good
#> 1:     99.6

# Flagged readings with their anomaly codes
dt[qc_flag > 1L, .(timestamp, raw_value, qc_value, qc_flag, qc_y_code)]
#>                timestamp raw_value qc_value qc_flag qc_y_code
#> 1: 2023-03-14 06:15:00        892       NA       4         8
#> 2: 2023-03-14 06:00:00        891       NA       4         8
#> ...
```

### 4.3 Comparing raw and QC-accepted values

```{r raw-vs-qc}
library(ggplot2)

dt <- read_silver("data/hydrometric", "39001", "Q",
                  start = "2023-03-01", end = "2023-04-01",
                  min_quality = NULL)@readings

ggplot(dt, aes(x = dateTime)) +
  geom_line(aes(y = raw_value), colour = "#AAAAAA", linewidth = 0.4) +
  geom_line(aes(y = qc_value),  colour = "#005EB8", linewidth = 0.8,
            na.rm = TRUE) +
  labs(
    title    = "Raw vs QC-accepted flow",
    subtitle = "Grey = raw Bronze value; Blue = qc_value (NA gaps = Rejected)",
    x        = NULL,
    y        = "Flow (m\u00b3/s)"
  ) +
  theme_bw()
```

---

## 5. Multiple gauges and catchment reads

### 5.1 Multiple gauges

`read_silver_multi()` reads several gauges in one call and returns a named
list of S7 objects. Gauges with no Silver data are skipped with a warning
rather than erroring.

```{r read-multi}
objs <- read_silver_multi(
  root      = "data/hydrometric",
  gauge_ids = c("39001", "39002", "39003"),
  data_type = "Q",
  start     = "2023-01-01",
  end       = "2023-12-31"
)

names(objs)         # gauges with data present
#> [1] "39001" "39002"   # 39003 had no Silver data — skipped with warning

objs[["39001"]]@n_rows
objs[["39002"]]@n_rows
```

### 5.2 All gauges in a catchment

`read_silver_catchment()` looks up gauge IDs from a registry and delegates
to `read_silver_multi()`:

```{r read-catchment}
reg  <- build_gauge_registry("data/gauge_registry.csv")

objs <- read_silver_catchment(
  root         = "data/hydrometric",
  catchment_id = "Thames_Upper",
  data_type    = "Q",
  registry     = reg,
  start        = "2023-01-01",
  end          = "2023-12-31",
  min_quality  = 3L   # Good + Suspect
)

# Row counts per gauge
lapply(objs, function(x) x@n_rows)
```

### 5.3 Combining catchment gauges into a long table

```{r combine-catchment}
long_dt <- data.table::rbindlist(lapply(objs, as_long), fill = TRUE)

# Daily mean per gauge
daily <- long_dt[, .(mean_flow = mean(value, na.rm = TRUE)), by = .(site_id, date)]

ggplot(daily, aes(x = date, y = mean_flow, colour = site_id)) +
  geom_line(linewidth = 0.6) +
  labs(title = "Daily mean flow — Thames Upper catchment",
       x = NULL, y = "Flow (m\u00b3/s)", colour = "Gauge") +
  theme_bw()
```

---

## 6. Silver catalogue

The catalogue functions let you query the Silver store without reading data.

```{r silver-catalogue}
# All gauges with Silver data
list_silver_gauges(root = "data/hydrometric")
#>    site_id  category     data_type  year_min year_max file_count
#> 1: 39001    hydrometric  Q          2020     2024        5
#> 2: 39002    hydrometric  Q          2021     2024        4
#> 3: 39001    hydrometric  P          2020     2024        5

# QC flag breakdown per gauge
summarise_silver_coverage(root = "data/hydrometric", data_type = "Q")
#>    site_id  data_type  start       end         n_rows  n_good n_estimated
#> 1: 39001    Q          2020-01-01  2024-12-31  175200  174850           0
#>    n_suspect n_rejected pct_good
#>          310         40     99.8

# Specific gauge
summarise_silver_coverage(root = "data/hydrometric", site_id = "39001")
```

### 6.1 Gap detection

```{r silver-gaps}
check_silver_gaps(
  root      = "data/hydrometric",
  site_id   = "39001",
  data_type = "Q"
)
#> Gap analysis (Silver): 39001 | Q
#> Expected interval: 15min
#> ─────────────────────────────────────────────────────────────────────
#>    gap_start               gap_end                 n_missing
#> 1: 2021-08-12 14:00:00 UTC 2021-08-12 14:45:00 UTC  4
#> 2: 2022-11-03 09:00:00 UTC 2022-11-03 10:45:00 UTC  8
```

---

## 7. Silver file paths

`silver_path()` builds the canonical Silver Parquet file path for a dataset,
following the same folder structure as Bronze:

```
silver/<category>/<data_type>/<year>/<dataset_id>.parquet
```

```{r silver-path}
silver_path(
  output_dir   = "data/hydrometric",
  category     = "hydrometric",
  data_type    = "Q",
  dataset_id   = "EA_39001_Q_20260115"
)
#> [1] "data/hydrometric/silver/hydrometric/Q/2026/EA_39001_Q_20260115.parquet"
```

---

## Summary of key functions

| Task | Function |
|---|---|
| Promote Bronze → Silver | `promote_to_silver(bronze_data, output_dir, ...)` |
| Stage–flow consistency check | `qc_h_stage_flow_consistency(stage, flow)` |
| Rainfall dry-run check | `qc_p_dry_run(value, wet_period_mask)` |
| Spatial consistency check | `qc_p_temporal_consistency(value, neighbours)` |
| Read Silver data | `read_silver(root, gauge_id, data_type, start, end, min_quality)` |
| Read multiple gauges | `read_silver_multi(root, gauge_ids, data_type, ...)` |
| Read by catchment | `read_silver_catchment(root, catchment_id, data_type, registry, ...)` |
| List Silver gauges | `list_silver_gauges(root)` |
| QC coverage summary | `summarise_silver_coverage(root, data_type, site_id)` |
| Find gaps | `check_silver_gaps(root, site_id, data_type)` |
| Build file path | `silver_path(output_dir, category, data_type, dataset_id)` |
